Supplementary Information:\\Evidence for black-swan events in animal populations

### 2015-11-24 text/supp.tex
Update before reformat

\title{Supporting Information:\\Evidence for black-swan events in animal
  populations}

\author{
Sean C. Anderson$^{1,2}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

@@

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error. The specific code for
the various models in our analysis is available at
\url{https://github.com/seananderson/heavy-tails}.

%\begin{spacing}{1.15}
\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=3> N;              // number of observations
  vector[N] y;                 // vector to hold ln abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
  real<lower=-1, upper=1> phi; // AR1 parameter
  vector[N] U;                 // unobserved states
  real<lower=0> sigma_obs;     // specified observation error SD
}
transformed parameters {
  vector[N] epsilon;           // error terms
  epsilon[1] <- 0;
  for (i in 2:N) {
    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
                       - (phi * epsilon[i - 1]);
  }
}
model {
  // priors:
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  phi ~ normal(0, 1);
  // data model:
  for (i in 2:N) {
    U[i] ~ student_t(nu,
                     lambda + b * U[i - 1]
                     + phi * epsilon[i - 1],
                     sigma_proc);
  }
  y ~ normal(U, sigma_obs);
}
\end{verbatim}
\end{footnotesize}

>>>>

\title{Supplementary Information:\\Evidence for black-swan events in animal
  populations}

\author{
Sean C. Anderson$^{1,2}$ \and
Trevor A. Branch$^3$ \and
Andrew B. Cooper$^2$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

We separated our simulation into these two components to avoid confounding two
issues. (1) With smaller sample sizes, there may not be a stochastic draw from
the tails of a distribution. In that case, no model, no matter how perfect,
will be able to detect the shape of the tails. (2) Complex models may return
biased parameter estimates if there are conceptual, computational, or coding
errors. Our first simulation tested the first issue and our second simulation
tested the latter. In general, our simulations show that, if anything, our
model under predicts the magnitude and probability of heavy tailed
events---especially given the length of the time series in the GPDD.

First, we tested the ability to estimate $\nu$ given different true values of
$\nu$ and different sample sizes. We took stochastic draws from t
distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
[$\approx$ normal]), with central tendency parameters of $0$, and scale
parameters of $1$. We started with $1600$ stochastic draws and then fit the
models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
we recorded the posterior samples of $\nu$.

Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates when the process noise was chosen so that
appropriate tail events were present. To generate these process deviations for
the $\nu = 3$ and $\nu = 5$ scenarios, we repeatedly drew proposed candidate
process deviations and estimated the central tendency, scale, and $\nu$ values
each time. We recorded when $\hat{\nu}$ (median of the posterior) was within
$0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this
set of random seed values in our Gompertz simulation.

@@

We then fit our Gompertz models to the simulated datasets with all parameters
(except $\nu$) set near the median values estimated in the GPDD. We repeated
this with $50$ and $100$ samples without observation error, $50$ samples with
observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
same observation error and a Gompertz model that allowed for correctly
specified observation error magnitude. Our results indicate that the Gompertz
model can recapture the true value of $\nu$ when the process noise was chosen
so that appropriate tail events were present (Extended Data Fig.
5 upper panels). The addition of observation error caused the
model to tend to underestimate the degree of heavy-tailedness. Fitting a model
with correctly specified observation error did not make substantial
improvements to model bias (Extended Data Fig. 5).

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
models distinguished heavy and not-heavy tails reasonably well (Extended Data
Fig. 5 lower panels). Without observation error, and using a
probability of $0.5$ as a threshold, the model correctly classified all
simulated systems with normally distributed process noise as not heavy tailed.
The model would have miscategorized only one of $40$ simulations at $\nu = 5$
across simulated populations with $50$ or $100$ time steps (Extended Data Fig.
5, scenarios 1 and 2 in lower row, second panel from left).
The model would have correctly categorized all cases where the process noise
was not heavy tailed (Extended Data Fig. 5 bottom-right
panel) and all cases where $\nu = 3$ and there was not observation error. With
$0.2$ standard deviations of observation error, the model still categorized
\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
case when $\nu = 3$. Allowing for observation error made little improvement to
the detection of heavy tails. Therefore, we chose to focus on the simpler
model without observation error in the main text, particularly given that the
true magnitude of observation error was unknown in the empirical data.

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model. The specific code for
the various models in our analysis is available at
\url{https://github.com/seananderson/heavy-tails} (currently private).

%\begin{spacing}{1.15}
%% \begin{footnotesize}
%% \begin{verbatim}
%% data {
%%   int<lower=3> N;              // number of observations
%%   vector[N] y;                 // vector to hold ln abundance observations
%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
%% }
%% parameters {
%%   real lambda;                 // Gompertz growth rate parameter
%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
%%   real<lower=0> sigma_proc;    // process noise scale parameter
%%   real<lower=2> nu;            // t-distribution degrees of freedom
%%   real<lower=-1, upper=1> phi; // AR1 parameter
%%   vector[N] U;                 // unobserved states
%%   real<lower=0> sigma_obs;     // specified observation error SD
%% }
%% transformed parameters {
%%   vector[N] epsilon;           // error terms
%%   epsilon[1] <- 0;
%%   for (i in 2:N) {
%%     epsilon[i] <- U[i] - (lambda + b * U[i - 1])
%%                        - (phi * epsilon[i - 1]);
%%   }
%% }
%% model {
%%   // priors:
%%   nu ~ exponential(nu_rate);
%%   lambda ~ normal(0, 10);
%%   sigma_proc ~ cauchy(0, 2.5);
%%   phi ~ normal(0, 1);
%%   // data model:
%%   for (i in 2:N) {
%%     U[i] ~ student_t(nu,
%%                      lambda + b * U[i - 1]
%%                      + phi * epsilon[i - 1],
%%                      sigma_proc);
%%   }
%%   y ~ normal(U, sigma_obs);
%% }
%% \end{verbatim}
%% \end{footnotesize}
%%

\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=1> N;              // number of observations
  vector[N] y;                 // vector to hold log abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
}
model {
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  for (i in 2:N) {
    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
  }
}
\end{verbatim}
\end{footnotesize}
